\documentclass{article}
\usepackage[a4paper,total={170mm,257mm},left=15mm,right=15mm, top=20mm,bottom=25mm]{geometry}
\usepackage{amssymb}
\begin{document}
\begin{center}
	\LARGE \textbf{Network Security}\\
	\vspace{20pt}
\end{center}
\large
\begin{itemize}
	\item Sniffing programs: programs which can analyze the headers in a packet, for info like sender IP, receiver IP, sequence numbers etc.
	\begin{itemize}
		\item Active sniffing: Sniffer acquires the packet because it is in the same network. Done only on broadcasted packets.
		\item Passive sniffing: Packets are acquired by changing the default gateway of sender so that the packets get routed through the sniffer. Gateway is changed by \textit{ARP spoofing}.
	\end{itemize}
	\item ARP (Address Resolution Protocol): \begin{itemize}
		\item Process of getting physical/MAC address from logical/IP address.
		\item A request is broadcasted like ``host with IP of x.x.x.x, send your MAC address". Every host in the network checks if the IP matches with it's own, and if it does, it sends a unicast with it's MAC address.
		\item The MAC address is at that moment. It may change with time.
		\item IP vs MAC address table is maintained in ARP cache, and it is cleared after some time.
	\end{itemize}
	\item RARP: Reverse ARP. Gets IP from MAC.
	\item ARP Spoofing:\begin{itemize}
		\item ARP cache is updated whether an ARP request is sent or not. If a spoofed ARP reply is sent, it will cause hosts to update the cache with wrong entry. This is called cache poisoning.
		\item If a host $C$ updates host $A$'s cache so that it's cache contains entry of $IP_B$,$MAC_C$, and updates cache of $B$ to contain entry $IP_A$,$MAC_C$, all messages that $A$ and $B$ send to eachother will go through $C$ first. $C$ can then alter the messages before forwarding to the meant destination.
		\item If a single IP is poisoned from all hosts with some MAC address that does not exist, host with that IP will not receive any messages. This is a type of DoS attack.
	\end{itemize}
\end{itemize}
war driving:can I find a wireless network?
war dialing:can I find a modem to connect to?
SQL injection
cross site scripting

\end{document}


